The LBRY Trending Algorithm

### 2021-11-17 tex/ms.tex
touches

In 2021, Jack Robison rewrote a lot of the hub code to improve performance
and to ensure hub data integrity when a chain reorganisation occurs. The
backend now uses leveldb to store transaction and claim information, and
Elasticsearch to store claim information in a suitable form for clients to
query. Leveldb will be replaced with rocksdb soon.

@@

The formula for updating a trending score is Equation~\ref{eqn:decay_with_spike}:
\begin{align}
Y_{t+1} &= kY_t + M(\ell_t, \ell_{t+1}).
\end{align}
However, instead of storing $Y$ for each claim, we could instead store
a transformed version of it:
\begin{align}
Z_t &= Y_t k^{-t}.
\end{align}
The updating formula for $Y$ can then be rewritten as
\begin{align}
Z_{t+1} k^{t+1} &= kZ_t k^t + M(\ell_t, \ell_{t+1}).
\end{align}
Solving for $Z_{t+1}$ gives
\begin{align}
Z_{t+1} &= kZ_t k^t k^{-(t+1)} + M(\ell_t, \ell_{t+1})k^{-(t+1)} \\
        &= Z_t + M(\ell_t, \ell_{t+1})k^{-(t+1)}.
\end{align}
Thus, if we maintain $Z$ instead of $Y$, 
we replace `exponential decay' with maintenance of a constant trending
score for all untouched claims. As time goes on, the mass of applied spikes
merely grows exponentially (recall that $k$ is just below 1, so $k$
raised to a negative power becomes large and positive),
and that is how recent changes become more significant than old ones.

The only difficulty now is how to add exponentially-large values to touched
claims while avoiding overflow.

>>>>

In 2021, Jack Robison rewrote a lot of the hub code to improve performance
and to ensure hub data integrity when a chain reorganisation occurs. The
backend now uses leveldb to store transaction and claim information, and
Elasticsearch to store claim information in a suitable form for clients to
query. Leveldb will be replaced with rocksdb soon.
As part of these changes, trending needed to be made more efficient. This
meant we made some modifications to it, some of which are pretty neat
in my opinion.

@@

The formula for updating a trending score is Equation~\ref{eqn:decay_with_spike}:
\begin{align}
Y_{t+1} &= kY_t + M(\ell_t, \ell_{t+1}).
\end{align}
However, instead of storing $Y$ for each claim, we could instead store
a transformed version of it:
\begin{align}
Z_t &= Y_t k^{-t}.
\end{align}
The updating formula for $Y$ can then be rewritten as
\begin{align}
Z_{t+1} k^{t+1} &= kZ_t k^t + M(\ell_t, \ell_{t+1}).
\end{align}
Solving for $Z_{t+1}$ gives
\begin{align}
Z_{t+1} &= kZ_t k^t k^{-(t+1)} + M(\ell_t, \ell_{t+1})k^{-(t+1)} \\
        &= Z_t + M(\ell_t, \ell_{t+1})k^{-(t+1)}.
\end{align}
Thus, if we maintain $Z$ instead of $Y$, 
we replace `exponential decay' with maintenance of a constant trending
score for all untouched claims --- the formula reduces to $Z_{t+1}=Z_t$.
As time goes on, the mass of applied spikes
merely grows exponentially (recall that $k$ is just below 1, so $k$
raised to a negative power becomes large and positive),
and that is how recent changes become more significant than old ones.

The only difficulty now is how to add exponentially-large values to touched
claims while avoiding overflow. This requires another nonlinear transformation
which I informally call `squashing'.
